---
title: "R Notebook"
output: html_notebook
---

If you opened this rmd directly, the file paths will not work, be sure you opened the .rproj file and then opened the this rmd within rstudio

Leave coverages in current folders or code won't work!

```{r}
library(tidyverse)
library(tmap)
library(terra) #the raster package is going away, switching to terra
library(whitebox)
library(RSAGA)

#install_whitebox()
wbt_init()
```
Whitebox tools manual: https://www.whiteboxgeo.com/manual/wbt_book/intro.html
RSAGA: https://cran.r-project.org/web/packages/RSAGA/RSAGA.pdf

Prep DEM

Feature preserving smooth
Fill single cell pits
Breach larger depressions

```{r}
#read regular and hydro prepped 1m DEMs
dem <- rast("dem1m/dem1m.tif")
hydrodem <- rast("hydem1m.tif/hydem1m.tif")

#generate 5m dem
terra::aggregate(dem,
                 filename = "dem_5m_aggregated.tif",
                 fact = 5,
                 fun = "mean")

#generate hydro prepped using hydem1m
terra::aggregate(hydrodem,
                 filename = "hydrodem_5m_aggregated.tif",
                 fact = 5,
                 fun = "mean")

wbt_feature_preserving_smoothing(
  dem ="hydrodem_5m_aggregated.tif",
  output = "hydem5m_smooth.tif",
  filter = 5,
  norm_diff = 15,
  num_iter = 3,
  max_diff = 0.5)

wbt_fill_single_cell_pits(
  dem = "hydem5m_smooth.tif",
  output = "hydem5m_smooth_filled.tif")

wbt_breach_depressions_least_cost(
  dem = "hydem5m_smooth_filled.tif",
  output = "hydem5m_smooth_filled_breached.tif",
  dist = 5,
  fill = TRUE)

wbt_hillshade(
  dem = "hydem5m_smooth_filled_breached.tif",
  output = "hydem5m_prepped_hillshade.tif")
```
Look at hillshade of hydro prepped DEM
```{r}
hillshade <- rast("hydem5m_prepped_hillshade.tif")

tmap_mode("view")
tm_shape(hillshade)+
  tm_raster(style = "cont",palette = "-Greys", legend.show = FALSE)+
  tm_scale_bar()
```

Original topo metrics (Gillin et al 2015)
        TPI 100
        TWId
        EDb
        Uaab
New metrics moving forward in Whitebox or RSAGA
        Max slope (1m) (use 1m DEM)
        TPI 20, 100, 500 (SAGA, DevFromMeanElev in WBT?) (5m but not smoothed) 
        Multi-resolution valley bottom (MRVB; SAGA; 5m not smoothed)
        TWId (GRASS, raster calculator; 5m, hydro prepped)
            Feature preserving filter DEM (GRASS)
            Ln(UAA)/tan(downslope index 5m distance) SAGA
        #downslope unsaturated flowpath with WBT free trial (1m, 5m, smooth/not)

TPI 100
https://www.whiteboxgeo.com/manual/wbt_book/available_tools/geomorphometric_analysis.html?highlight=topographic%20position#relativetopographicposition

filter size is in # of cells
```{r}
#TPI 100
wbt_dev_from_mean_elev(
  dem = "dem_5m_aggregated.tif", 
  output = "tpi100m.tif", 
  filterx = 100, 
  filtery = 100)

#slope from unprocessed 1m DEM
wbt_slope(
  dem = "dem1m/dem1m.tif", 
  output = "slope.tif")

#log d8 flow accumulation
wbt_fd8_flow_accumulation(
  dem = "hydem5m_smooth_filled_breached.tif",
  output = "hydem5m_d8fa.tif",
  out_type = 'catchment area',
  log = TRUE
)

#tangent downslope index
wbt_downslope_index(
  dem = "hydem5m_smooth_filled_breached.tif",
  output= "hydem5m_downslope_index.tif",
  out_type = 'tangent'
)

#read flow accumulation and downslope index for raster calc operation
d8fa <- rast("hydem5m_d8fa.tif")
tan_ds_index <- rast("hydem5m_downslope_index.tif")

#divide log flow accumulation by tangend of downslope index
TWId <- d8fa / tan_ds_index

#output TWId to file
writeRaster(TWId, "hydem5m_TWId.tif")
```
MRVB
To run this you have to link to a SAGA installation on your computer. See more here: https://geocompr.robinlovelace.net/gis.html?q=saga#rsaga

Usage: saga_cmd ta_morphometry 8 [-DEM <str>] [-MRVBF <str>] [-MRRTF <str>] [-T_SLOPE <double>] [-T_PCTL_V <double>] [-T_PCTL_R <double>] [-P_SLOPE <double>] [-P_PCTL <double>] [-UPDATE <str>] [-CLASSIFY <str>] [-MAX_RES <double>]
  -DEM:<str>        	Elevation
	Grid, input
  -MRVBF:<str>      	MRVBF
	Grid, output
  -MRRTF:<str>      	MRRTF
	Grid, output
  -T_SLOPE:<double> 	Initial Threshold for Slope
	Floating point
	Minimum: 0.000000
	Maximum: 100.000000
	Default: 16.000000
  -T_PCTL_V:<double>	Threshold for Elevation Percentile (Lowness)
	Floating point
	Minimum: 0.000000
	Maximum: 1.000000
	Default: 0.400000
  -T_PCTL_R:<double>	Threshold for Elevation Percentile (Upness)
	Floating point
	Minimum: 0.000000
	Maximum: 1.000000
	Default: 0.350000
  -P_SLOPE:<double> 	Shape Parameter for Slope
	Floating point
	Default: 4.000000
  -P_PCTL:<double>  	Shape Parameter for Elevation Percentile
	Floating point
	Default: 3.000000
  -UPDATE:<str>     	Update Views
	Boolean
	Default: 1
  -CLASSIFY:<str>   	Classify
	Boolean
	Default: 0
  -MAX_RES:<double> 	Maximum Resolution (Percentage)
	Floating point
	Minimum: 0.000000
	Maximum: 100.000000
	
Default values for MRVBF run:
Elevation: dem_saga
MRVBF: MRVBF
MRRTF: MRRTF
Initial Threshold for Slope: 16
Threshold for Elevation Percentile (Lowness): 0.4
Threshold for Elevation Percentile (Upness): 0.35
Shape Parameter for Slope: 4
Shape Parameter for Elevation Percentile: 3
Update Views: true
Classify: false
Maximum Resolution (Percentage): 100
```{r}
sagaenv <- rsaga.env("//Mac/Home/Documents/saga-8.1.1_x64/saga-8.1.1_x64")

#rsaga.import.gdal(in.grid = "dem_5m_aggregated.tif",
  #                out.grid = "dem_saga",
  #            env = sagaenv)

rast("dem_5m_aggregated.tif") %>% 
  terra::writeRaster(filename = "dem_5m_aggregated.sdat",
              filetype = "SAGA")

rsaga.get.modules(libs = "ta_morphometry",
                  env = sagaenv)

#rsaga.get.usage(lib = "ta_morphometry", 
#                module = "Multiresolution Index of Valley Bottom Flatness (MRVBF)", 
#                env = sagaenv)

params = list(DEM = "dem_5m_aggregated.sgrd",
              MRVBF = "mrvbf.sdat",
              MRRTF = "mrrtf.sdat")

rsaga.geoprocessor(lib = "ta_morphometry", 
                   module = "Multiresolution Index of Valley Bottom Flatness (MRVBF)", 
                   param = params,
                   env = sagaenv)

mrvbf <- rast("mrvbf.sdat")
crs(mrvbf) <- "epsg:26919"
writeRaster(mrvbf, "mrvbf_projected.tif")
```

look at outputs
```{r}
toview <- raster("tpi100m.tif")

tmap_mode("view")
tm_shape(toview)+
  tm_raster(style = "cont", legend.show = FALSE)+
  tm_scale_bar()
```

