---
title: "R Notebook"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

If you opened this rmd directly, the file paths will not work, be sure
you opened the .rproj file and then opened the this rmd within rstudio

Leave coverages in current folders or code won't work!

```{r}
library(tidyverse)
library(tmap)
library(terra) #the raster package is going away, switching to terra
library(raster)
library(whitebox)
library(RSAGA)
library(sf)

#install_whitebox()
wbt_init()
```

Whitebox tools manual:
<https://www.whiteboxgeo.com/manual/wbt_book/intro.html> RSAGA:
<https://cran.r-project.org/web/packages/RSAGA/RSAGA.pdf>

Prep DEM for hydro related metrics and make 5m regular dem

Feature preserving smooth Fill single cell pits Breach larger
depressions

```{r}
#Aggregate 1 m DEM to 5 m
wbt_aggregate_raster(
  input = "knb-lter-hbr.211.2/dem1m.tif",
  output = "dem5m_noproj.tif",
  agg_factor = 5
  )

temp <- rast("dem5m_noproj.tif")
crs(temp) <- "epsg:26919"
writeRaster(temp, 
           "dem5m.tif",
           overwrite = TRUE)

#this strips the projection
#must add the projection back after
#calculating downslope index
#downslope index tool does not like 
#rasters output by terra functions
wbt_aggregate_raster(
  input = "knb-lter-hbr.211.2/hydem1mlpns.tif",
  output = "hydem5m.tif",
  agg_factor = 5
  )

#smooth 5m DEM
wbt_feature_preserving_smoothing(
  dem = "hydem5m.tif",
  output = "hydem5m_smooth.tif",
  filter = 11,
  norm_diff = 15,
  num_iter = 3,
  max_diff = 0.5)

#breach depressions
wbt_breach_depressions_least_cost(
  dem = "hydem5m_smooth.tif",
  output = "hydem5m_smooth_breached.tif",
  dist = 100,
  flat_increment = 0.01)

#fill remaining depressions
#there is a bug in the above function
#that leaves some depressions, 
#this fixes that
wbt_fill_depressions_wang_and_liu(
  dem = "hydem5m_smooth_breached.tif",
  output = "hydem5m_smooth_breached_filled.tif"
)

```

Code used to check on DEMs generated above
```{r}
wbt_hillshade(
  dem = "hydem5m_smooth_filled_breached2.tif",
  output = "hydem5m_prepped_hillshade.tif")

hillshade <- rast("hydem5m_prepped_hillshade.tif")

tmap_mode("plot")
tm_shape(hillshade)+
  tm_raster(style = "cont",palette = "-Greys", legend.show = FALSE)+
  tm_scale_bar()
```

Original topo metrics (Gillin et al 2015) TPI 100 TWId EDb Uaab New
metrics moving forward in Whitebox or RSAGA Max slope (1m) (use 1m DEM)
TPI 20, 100, 500 (SAGA, DevFromMeanElev in WBT?) (5m but not smoothed)
Multi-resolution valley bottom (MRVB; SAGA; 5m not smoothed) TWId
(GRASS, raster calculator; 5m, hydro prepped) Feature preserving filter
DEM (GRASS) Ln(UAA/tan(downslope index 5m distance))? SAGA \#downslope
unsaturated flowpath with WBT free trial (1m, 5m, smooth/not)

TPI
<https://www.whiteboxgeo.com/manual/wbt_book/available_tools/geomorphometric_analysis.html?highlight=topographic%20position#relativetopographicposition>

```{r}
#TPI 20
wbt_dev_from_mean_elev(
  dem = "dem5m.tif", 
  output = "tpi20m.tif", 
  filterx = 20, 
  filtery = 20)

#TPI 100
wbt_dev_from_mean_elev(
  dem = "dem5m.tif", 
  output = "tpi100m.tif", 
  filterx = 100, 
  filtery = 100)

#TPI 500
wbt_dev_from_mean_elev(
  dem = "dem5m.tif", 
  output = "tpi500m.tif", 
  filterx = 500, 
  filtery = 500)

#slope from unprocessed 1m DEM
wbt_slope(
  dem = "knb-lter-hbr.211.2/dem1m.tif", 
  output = "slope.tif")

#aggregate slope to 5m
wbt_aggregate_raster(
  input = "slope.tif",
  output = "slope5m.tif",
  agg_factor = 5
  )

#log md inf (Seibert & McGlynn 07) flow accumulation
#1.65 seconds
wbt_md_inf_flow_accumulation(
  dem = "hydem5m_smooth_breached_filled.tif",
  output = "hydem5m_mdinf_fa.tif",
  out_type = "specific contributing area"
)

#tangent downslope index
wbt_downslope_index(
  dem = "hydem5m_smooth_breached_filled.tif",
  output= "hydem5m_downslope_index.tif",
  drop = 5,
  out_type = "tangent"
)

#read flow accumulation and downslope index for raster calc operation
mdinf <- rast("hydem5m_mdinf_fa.tif")
tan_ds_index <- rast( "hydem5m_downslope_index.tif")

#divide log flow accumulation by tangent of downslope index
TWId <- base::log(mdinf / tan_ds_index)
crs(TWId) <- "epsg:26919"

#output TWId to file
writeRaster(TWId, "hydem5m_TWId.tif", overwrite = TRUE)
```

MRVB To run this you have to link to a SAGA installation on your
computer. See more here:
<https://geocompr.robinlovelace.net/gis.html?q=saga#rsaga>

Usage: saga_cmd ta_morphometry 8 [-DEM <str>] [-MRVBF <str>] [-MRRTF
<str>] [-T_SLOPE <double>] [-T_PCTL_V <double>] [-T_PCTL_R <double>]
[-P_SLOPE <double>] [-P_PCTL <double>] [-UPDATE <str>] [-CLASSIFY <str>]
[-MAX_RES <double>] -DEM:<str> Elevation Grid, input -MRVBF:<str> MRVBF
Grid, output -MRRTF:<str> MRRTF Grid, output -T_SLOPE:<double> Initial
Threshold for Slope Floating point Minimum: 0.000000 Maximum: 100.000000
Default: 16.000000 -T_PCTL_V:<double> Threshold for Elevation Percentile
(Lowness) Floating point Minimum: 0.000000 Maximum: 1.000000 Default:
0.400000 -T_PCTL_R:<double> Threshold for Elevation Percentile (Upness)
Floating point Minimum: 0.000000 Maximum: 1.000000 Default: 0.350000
-P_SLOPE:<double> Shape Parameter for Slope Floating point Default:
4.000000 -P_PCTL:<double> Shape Parameter for Elevation Percentile
Floating point Default: 3.000000 -UPDATE:<str> Update Views Boolean
Default: 1 -CLASSIFY:<str> Classify Boolean Default: 0 -MAX_RES:<double>
Maximum Resolution (Percentage) Floating point Minimum: 0.000000
Maximum: 100.000000

Default values for MRVBF run: Elevation: dem_saga MRVBF: MRVBF MRRTF:
MRRTF Initial Threshold for Slope: 16 Threshold for Elevation Percentile
(Lowness): 0.4 Threshold for Elevation Percentile (Upness): 0.35 Shape
Parameter for Slope: 4 Shape Parameter for Elevation Percentile: 3
Update Views: true Classify: false Maximum Resolution (Percentage): 100

```{r}
#tell R where saga is
sagaenv <- rsaga.env("//Mac/Home/Documents/saga-8.1.1_x64/saga-8.1.1_x64")

#convert 5m raster to saga file format
rast("dem_5m_aggregated.tif") %>% 
  terra::writeRaster(filename = "dem_5m_aggregated.sdat",
              filetype = "SAGA")

rsaga.get.modules(libs = "ta_morphometry",
                  env = sagaenv)

#rsaga.get.usage(lib = "ta_morphometry", 
#                module = "Multiresolution Index of Valley Bottom Flatness (MRVBF)", 
#                env = sagaenv)

#set parameters for mrvbf run (additional available are above this chunk)
params = list(DEM = "dem_5m_aggregated.sgrd",
              MRVBF = "mrvbf.sdat",
              MRRTF = "mrrtf.sdat")

#run mrvbf using parameters aboge
rsaga.geoprocessor(lib = "ta_morphometry", 
                   module = "Multiresolution Index of Valley Bottom Flatness (MRVBF)", 
                   param = params,
                   env = sagaenv)

#read in the result of above, assive projection, write back out projected
mrvbf <- rast("mrvbf.sdat")
crs(mrvbf) <- "epsg:26919"
writeRaster(mrvbf, "mrvbf_projected.tif")
```

The code below reads in the coverages generated above, stacks them into
one big multi layer grid, and then extracts the values form each at the
point locations of soil observations

```{r}
#read 5 m rasters
all_layers <- rast(c(
  "slope_5m_aggregated.tif",
  "dem_5m_aggregated.tif",
  "hydem5m_TWId.tif",
  "mrvbf_projected.tif",
  "tpi20m.tif",
  "tpi100m.tif",
  "tpi500m.tif"
))

#read 1 m rasters
slope1m <- rast("slope.tif")

#read point shapefile
observation_locations <- vect("HBEF_NEW_biscuit_Pedon/HBEF_NEW_biscuit_Pedon.shp")

#extract values at points from 5m raster layers, drop ID and rename slope bc those exist already in the imported shapefile
raster_values <- extract(x = all_layers, #raster 
                  y = observation_locations, #points
                  method = "simple") %>%
                select(-ID) %>%
                dplyr::rename(slope5m = slope)

#have to extract 1 m slope separately bc of different resolution
slope_1m_values <- extract(x = slope1m, #raster 
                  y = observation_locations, #points
                  method = "simple")

#add extracted values to shapefile
observation_locations <- cbind(observation_locations, raster_values, slope_1m_values)

#write a new shapefile with extracted values
writeVector(observation_locations,
            "HBEF_NEW_biscuit_Pedon/HBEF_NEW_biscuit_Pedon_extracted_values.shp",
            filetype = "ESRI Shapefile")

#export csv of extracted values
as_tibble(observation_locations) %>%
  write_csv("biscuit_pedon_extracted_values.csv")
```

Create spatialpixelsdataframe

```{r}
library(raster)

  slope <- raster("slope_5m_aggregated.tif")
  dem <- raster("dem_5m_aggregated.tif")
  twid <- raster("hydem5m_TWId.tif")
  mrvbf <- raster("mrvbf_projected.tif")
  tpi20 <- raster("tpi20m.tif")
  tpi100 <- raster("tpi100m.tif")
  tpi500 <- raster("tpi500m.tif")
  
all_layers_raster <- stack(slope, dem, twid, mrvbf, tpi20, tpi100, tpi500)

spdf_all_layers <- as(all_layers_raster, "SpatialPixelsDataFrame")
```
