---
title: "R Notebook"
output: html_notebook
editor_options: 
  markdown: 
    wrap: 72
---

If you opened this rmd directly, the file paths will not work, be sure
you opened the .rproj file and then opened the this rmd within rstudio

Leave coverages in current folders or code won't work!

```{r echo = FALSE}
library(tidyverse)
library(tmap)
library(terra) #the raster package is going away, switching to terra
library(raster)
library(whitebox)
library(RSAGA)
library(sf)
library(qgisprocess)
library(caret)

algos <- qgis_algorithms()
#install_whitebox()
wbt_init()
```

Whitebox tools manual:
<https://www.whiteboxgeo.com/manual/wbt_book/intro.html> RSAGA:
<https://cran.r-project.org/web/packages/RSAGA/RSAGA.pdf>

Prep DEM for hydro related metrics and make 5m regular dem

Feature preserving smooth Fill single cell pits Breach larger
depressions

```{r}
#originaldem <- "knb-lter-hbr.211.2/dem1m.tif"
#hydro_prepped_orig_dem <- "knb-lter-hbr.211.2/hydem1mlpns.tif"

#data_dir <- "ConePond/"
#originaldem <- "cnp1mdem.tif"
#hydro_prepped_orig_dem <- "cnp1mdem.tif"

#data_dir <- "BlackPond/" #with a / after ie. directory/
#originaldem <- "bk1mdem.tif"
#hydro_prepped_orig_dem <- "bk1mdem.tif"
#epsg_orig <- "epsg:26919"
#
data_dir <- "HBValley/" #with a / after ie. directory/
#originaldem <- "knb-lter-hbr.211.2/dem1m.tif"
#hydro_prepped_orig_dem <- "knb-lter-hbr.211.2/hydem1mlpns.tif"
#epsg_orig <- "epsg:26919"


#Aggregate 1 m DEM to 5 m
wbt_aggregate_raster(
  input = paste0(data_dir, originaldem),
  output = paste0(data_dir, "dem5m_noproj.tif"),
  agg_factor = 5
  )

#aggregate strips the projection, this adds it back in
temp <- rast(paste0(data_dir, "dem5m_noproj.tif"))
crs(temp) <- epsg_orig
writeRaster(temp, 
           paste0(data_dir, "dem5m.tif"),
           overwrite = TRUE)

#this strips the projection
#must add the projection back after
#calculating downslope index
#downslope index tool does not like 
#rasters output by terra functions
wbt_aggregate_raster(
  input = paste0(data_dir, hydro_prepped_orig_dem),
  output = paste0(data_dir, "hydem5m.tif"),
  agg_factor = 5
  )

#smooth 5m DEM
wbt_feature_preserving_smoothing(
  dem = paste0(data_dir, "hydem5m.tif"),
  output = paste0(data_dir, "hydem5m_smooth.tif"),
  filter = 11,
  norm_diff = 15,
  num_iter = 3,
  max_diff = 0.5)

#breach depressions
wbt_breach_depressions_least_cost(
  dem = paste0(data_dir, "hydem5m_smooth.tif"),
  output = paste0(data_dir, "hydem5m_smooth_breached.tif"),
  dist = 100,
  flat_increment = 0.01)

#fill remaining depressions there is 
#a bug in the breach depressions least cost
#function that leaves some depressions, 
#this fixes that
wbt_fill_depressions_wang_and_liu(
  dem = paste0(data_dir, "hydem5m_smooth_breached.tif"),
  output = paste0(data_dir, "hydem5m_smooth_breached_filled.tif")
)

```

Code used to check on DEMs generated above
```{r}
wbt_hillshade(
  dem = paste0(data_dir, "hydem5m_smooth_breached_filled.tif"),
  output = paste0(data_dir, "hydem5m_prepped_hillshade.tif"))

hillshade <- rast(paste0(data_dir, "hydem5m_prepped_hillshade.tif"))

tmap_mode("plot")
tm_shape(hillshade)+
  tm_raster(style = "cont",palette = "-Greys", legend.show = FALSE)+
  tm_scale_bar()
```

Original topo metrics (Gillin et al 2015) TPI 100 TWId EDb Uaab 

New metrics moving forward in Whitebox or RSAGA Max slope (1m) (use 1m DEM)

TPI 20, 100, 500 (SAGA, DevFromMeanElev in WBT?) (5m but not smoothed)
Multi-resolution valley bottom (MRVB; SAGA; 5m not smoothed) TWId
(GRASS, raster calculator; 5m, hydro prepped) Feature preserving filter

DEM (GRASS) Ln(UAA/tan(downslope index 5m distance))? SAGA \#downslope
unsaturated flowpath with WBT free trial (1m, 5m, smooth/not)

TPI
<https://www.whiteboxgeo.com/manual/wbt_book/available_tools/geomorphometric_analysis.html?highlight=topographic%20position#relativetopographicposition>

```{r}
#TPI 20
#filter x and y dimensions below are in cells
#3 cells = 15 meters
##is 15 meters the RADIUS? RADIUS BELOW
wbt_dev_from_mean_elev(
  dem = paste0(data_dir, "dem5m.tif"), 
  output = paste0(data_dir, "tpi15m.tif"), 
  filterx = 3, 
  filtery = 3)

#TPI20
wbt_dev_from_mean_elev(
  dem = paste0(data_dir, "dem5m.tif"), 
  output = paste0(data_dir, "tpi20m.tif"), 
  filterx = 4, 
  filtery = 4)

#TPI 100
wbt_dev_from_mean_elev(
  dem = paste0(data_dir, "dem5m.tif"), 
  output = paste0(data_dir, "tpi100m.tif"), 
  filterx = 20, 
  filtery = 20)

#TPI 200 #CHANGED TO RADIUS OF 200
wbt_dev_from_mean_elev(
  dem = paste0(data_dir, "dem5m.tif"), 
  output = paste0(data_dir, "tpi200m.tif"), 
  filterx = 40, 
  filtery = 40)

#TPI 250
wbt_dev_from_mean_elev(
  dem = paste0(data_dir, "dem5m.tif"), 
  output = paste0(data_dir, "tpi250m.tif"), 
  filterx = 50, 
  filtery = 50)

#TPI 500
wbt_dev_from_mean_elev(
  dem = paste0(data_dir, "dem5m.tif"), 
  output = paste0(data_dir, "tpi500m.tif"), 
  filterx = 100, 
  filtery = 100)

#TPI 1000
wbt_dev_from_mean_elev(
  dem = paste0(data_dir, "dem5m.tif"), 
  output = paste0(data_dir, "tpi1000m.tif"), 
  filterx = 200, 
  filtery = 200)

#TPI 1500
wbt_dev_from_mean_elev(
  dem = paste0(data_dir, "dem5m.tif"), 
  output = paste0(data_dir, "tpi1500m.tif"), 
  filterx = 300, 
  filtery = 300)

#TPI 2000
wbt_dev_from_mean_elev(
  dem = paste0(data_dir, "dem5m.tif"), 
  output = paste0(data_dir, "tpi2000m.tif"), 
  filterx = 400, 
  filtery = 400)

wbt_slope(
  dem = paste0(data_dir, "dem5m.tif"), 
  output = paste0(data_dir, "slope_rad_5m.tif"),
  units = "radians")

##slope from unprocessed 1m DEM
#wbt_slope(
#  dem = paste0(data_dir, originaldem), 
#  output = paste0(data_dir, "slope.tif"))
#
#wbt_slope(
#  dem = paste0(data_dir, originaldem), 
#  output = paste0(data_dir, "slope_rad.tif"),
#  units = "radians")
#
##aggregate slope to 5m
#wbt_aggregate_raster(
#  input = paste0(data_dir, "slope.tif"),
#  output = paste0(data_dir, "slope5m_noproj.tif"),
#  agg_factor = 5
#  )
#
#wbt_aggregate_raster(
#  input = paste0(data_dir, "slope_rad.tif"),
#  output = paste0(data_dir, "slope5m_rad_noproj.tif"),
#  agg_factor = 5
#  )
#
##add projection to aggregated slope
#temp <- rast(paste0(data_dir, "slope5m_noproj.tif"))
#crs(temp) <- epsg_orig
#writeRaster(temp, 
#           paste0(data_dir, "slope5m.tif"),
#           overwrite = TRUE)
#
#temp <- rast(paste0(data_dir, "slope5m_rad_noproj.tif"))
#crs(temp) <- epsg_orig
#writeRaster(temp, 
#           paste0(data_dir, "slope5m_rad.tif"),
#           overwrite = TRUE)
#
#log md inf (Seibert & McGlynn 07) flow accumulation
#1.65 seconds
wbt_md_inf_flow_accumulation(
  dem = paste0(data_dir, "hydem5m_smooth_breached_filled.tif"),
  output = paste0(data_dir, "hydem5m_mdinf_fa.tif"),
  out_type = "specific contributing area"
)

#tangent downslope index
wbt_downslope_index(
  dem = paste0(data_dir, "hydem5m_smooth_breached_filled.tif"),
  output= paste0(data_dir, "hydem5m_downslope_index.tif"),
  drop = 5,
  out_type = "tangent"
)

#read flow accumulation and downslope index for raster calc operation
mdinf <- rast(paste0(data_dir, "hydem5m_mdinf_fa.tif"))
tan_ds_index <- rast(paste0(data_dir, "hydem5m_downslope_index.tif"))

#divide log flow accumulation by tangent of downslope index
TWId <- base::log(mdinf / tan_ds_index)
crs(TWId) <- epsg_orig

#output TWId to file
writeRaster(TWId, paste0(data_dir, "hydem5m_TWId.tif"), overwrite = TRUE)
```


Run BOSS model to predict bedrock
```{r}
source("boss.R")

boss(maxslope = raster(paste0(data_dir,"slope5m_rad.tif")),
     tpi15 = raster(paste0(data_dir,"tpi15m.tif")),
     tpi200 = raster(paste0(data_dir,"tpi200m.tif")),
     dir = data_dir)


```
Bedrock weighted UAA
```{r}
#euclidian distance from bedrock
boss01 <- raster(paste0(data_dir,"bossBRprediction_sagatpi.tif"))
boss01[boss01 == 1] <- 0
boss01[boss01 == 2] <- 1 #BR == 2 means bedrock in current prediction model

writeRaster(boss01, paste0(data_dir, "bossBRprediction_sagatpi.tif"), overwrite = TRUE)

wbt_euclidean_distance(
      input = paste0(data_dir,"bossBRprediction_sagatpi.tif"),
      output = paste0(data_dir, "EDb.tif"))

```
MRVB To run this you have to link to a SAGA installation on your
computer. See more here:
<https://geocompr.robinlovelace.net/gis.html?q=saga#rsaga>

Usage: saga_cmd ta_morphometry 8 [-DEM <str>] [-MRVBF <str>] [-MRRTF
<str>] [-T_SLOPE <double>] [-T_PCTL_V <double>] [-T_PCTL_R <double>]
[-P_SLOPE <double>] [-P_PCTL <double>] [-UPDATE <str>] [-CLASSIFY <str>]
[-MAX_RES <double>] -DEM:<str> Elevation Grid, input -MRVBF:<str> MRVBF
Grid, output -MRRTF:<str> MRRTF Grid, output -T_SLOPE:<double> Initial
Threshold for Slope Floating point Minimum: 0.000000 Maximum: 100.000000
Default: 16.000000 -T_PCTL_V:<double> Threshold for Elevation Percentile
(Lowness) Floating point Minimum: 0.000000 Maximum: 1.000000 Default:
0.400000 -T_PCTL_R:<double> Threshold for Elevation Percentile (Upness)
Floating point Minimum: 0.000000 Maximum: 1.000000 Default: 0.350000
-P_SLOPE:<double> Shape Parameter for Slope Floating point Default:
4.000000 -P_PCTL:<double> Shape Parameter for Elevation Percentile
Floating point Default: 3.000000 -UPDATE:<str> Update Views Boolean
Default: 1 -CLASSIFY:<str> Classify Boolean Default: 0 -MAX_RES:<double>
Maximum Resolution (Percentage) Floating point Minimum: 0.000000
Maximum: 100.000000

Default values for MRVBF run: Elevation: dem_saga MRVBF: MRVBF MRRTF:
MRRTF Initial Threshold for Slope: 16 Threshold for Elevation Percentile
(Lowness): 0.4 Threshold for Elevation Percentile (Upness): 0.35 Shape
Parameter for Slope: 4 Shape Parameter for Elevation Percentile: 3
Update Views: true Classify: false Maximum Resolution (Percentage): 100

```{r}
##tell R where saga is
#sagaenv <- rsaga.env("//Mac/Home/Documents/saga-8.1.1_x64/saga-8.1.1_x64")
#
##convert 5m raster to saga file format
#rast(paste0(data_dir, "dem5m.tif")) %>% 
#  terra::writeRaster(filename = paste0(data_dir, "dem5m.sdat"),
#              filetype = "SAGA")
#
#rsaga.get.modules(libs = "ta_morphometry",
#                  env = sagaenv)
#
##rsaga.get.usage(lib = "ta_morphometry", 
##                module = "Multiresolution Index of Valley Bottom Flatness (MRVBF)", 
##                env = sagaenv)
#
##set parameters for mrvbf run (additional available are above this chunk)
#params = list(DEM = paste0(data_dir, "dem5m.sgrd"),
#              MRVBF = paste0(data_dir, "mrvbf.sdat"),
#              MRRTF = paste0(data_dir, "mrrtf.sdat"))
#
##run mrvbf using parameters above
#rsaga.geoprocessor(lib = "ta_morphometry", 
#                   module = "Multiresolution Index of Valley Bottom Flatness (MRVBF)",
#                   param = params,
#                   env = sagaenv)
#
##read in the result of above, assign projection, write back out projected
#mrvbf <- rast(paste0(data_dir, "mrvbf.sdat"))
#crs(mrvbf) <- "epsg:26919"
#writeRaster(mrvbf, paste0(data_dir, "mrvbf_projected.tif"), overwrite = TRUE)


###MRVBF in QGIS package
dem <- raster(paste0(data_dir, "hydem5m.tif"))

#qgis_show_help("saga:multiresolutionindexofvalleybottomflatnessmrvbf")
qgis_run_algorithm(algorithm = "saga:multiresolutionindexofvalleybottomflatnessmrvbf",
                  DEM = dem,
                  MRVBF = paste0(data_dir, "mrvbf_qgis")
                   )

mrvbf <- rast(paste0(data_dir, "mrvbf_qgis.sdat"))
crs(mrvbf) <- "epsg:26919"
writeRaster(mrvbf, paste0(data_dir, "mrvbf.tif"), overwrite = TRUE)
```

Make UAAb bedrock weighted UAA using QGIS
```{r}
#prep BR weight grid by making BR values 10000
BR <- raster(paste0(data_dir,"bossBRprediction_sagatpi.tif"))
BR[BR == 2] <- 100000 #BR == 2 means bedrock in current prediction model
BR[BR == 1] <- 0
writeRaster(BR, paste0(data_dir, "BRweights.tif"), overwrite = TRUE)

#use QGIS to make UAAb
weights <- raster(paste0(data_dir,"BRweights.tif"))

qgis_run_algorithm(algorithm = "saga:catchmentarearecursive",
                   FLOW_UNIT = 0,
                  ELEVATION = dem,
                  METHOD = 3,
                  NO_NEGATIVES = 1,
                  FLOW = paste0(data_dir, "catcharea")
                   )

catcharea <- rast(paste0(data_dir, "catcharea.sdat"))
crs(catcharea) <- "epsg:26919"
writeRaster(catcharea, paste0(data_dir, "catcharea.tif"), overwrite = TRUE)

#weighted
#qgis_show_help("saga:catchmentarearecursive")
qgis_run_algorithm(algorithm = "saga:catchmentarearecursive",
                  ELEVATION = dem,
                  FLOW_UNIT = 0,
                  METHOD = 3,
                  WEIGHTS = weights,
                  NO_NEGATIVES = 1,
                  FLOW = paste0(data_dir, "catcharea_br")
                   )

catcharea_br <- rast(paste0(data_dir, "catcharea_br.sdat"))
crs(catcharea) <- "epsg:26919"
writeRaster(catcharea_br, paste0(data_dir, "catcharea_br.tif"), overwrite = TRUE)

diff <- catcharea_br/catcharea
normdiff <- (diff - minmax(diff)[1]) / (minmax(diff)[2] - minmax(diff)[1])
writeRaster(normdiff, paste0(data_dir, "uaab_norm2.tif"), overwrite = TRUE)

```
```{r}
#qgis_show_help("saga:topographicpositionindextpi")

qgis_run_algorithm(algorithm = "saga:topographicpositionindextpi",
                  DEM = dem,
                  RADIUS_MIN = 15,
                  TPI = paste0(data_dir, "tpi15saga")
                   )
tpi15saga <- rast(paste0(data_dir, "tpi15saga.sdat"))
crs(tpi15saga) <- "epsg:26919"
writeRaster(tpi15saga, paste0(data_dir, "tpi15saga.tif"), overwrite = TRUE)

qgis_run_algorithm(algorithm = "saga:topographicpositionindextpi",
                  DEM = dem,
                  RADIUS_MAX = 200,
                  RADIUS_MIN = 200,
                  TPI = paste0(data_dir, "tpi200saga")
                   )

tpi200saga <- rast(paste0(data_dir, "tpi200saga.sdat"))
crs(tpi200saga) <- "epsg:26919"
writeRaster(tpi200saga, paste0(data_dir, "tpi200saga.tif"), overwrite = TRUE)

```

