---
title: "R Notebook"
output: html_notebook
---
"/Users/jpgannon/Library/R/x86_64/4.1/library/whitebox/WBT/whitebox_tools "
```{r}
library(tidyverse)
library(tmap)
library(raster)
library(whitebox)
install_whitebox()
wbt_init()
```
Prep DEM

Feature preserving smooth
Fill single cell pits
Breach larger depressions

```{r}
#dem <- raster("hydem1m.tif/hydem1m.tif")

wbt_feature_preserving_smoothing(
  dem = "hydem1m.tif/hydem1m.tif",
  output = "hydem1m_smooth.tif",
  filter = 5,
  norm_diff = 15,
  num_iter = 3,
  max_diff = 0.5)

wbt_fill_single_cell_pits(
  dem = "hydem1m_smooth.tif",
  output = "hydem1m_smooth_filled.tif")

wbt_breach_depressions_least_cost(
  dem = "hydem1m_smooth_filled.tif",
  output = "hydem1m_smooth_filled_breached.tif",
  dist = 5,
  fill = TRUE)

wbt_hillshade(
  dem = "hydem1m_smooth_filled_breached.tif",
  output = "hydem1m_prepped_hillshade.tif")
```
Look at hillshade of prepped DEM
```{r}
hillshade <- raster("hydem1m_prepped_hillshade.tif")

tmap_mode("view")
tm_shape(hillshade)+
  tm_raster(style = "cont",palette = "-Greys", legend.show = FALSE)+
  tm_scale_bar()
```

Original topo metrics
        TPI 100
        TWId
        EDb
        Uaab
  New metrics moving forward in Whitebox or RSAGA
        Max slope (1m)
        TPI 20, 100, 500 (SAGA, 5m)
        Multi-resolution valley bottom (MRVB; SAGA; 5m)
        TWId (GRASS, raster calculator; 5m)
            Feature preserving filter DEM (GRASS)
            Ln(UAA)/tan(downslope index 5m distance) SAGA

TPI 100
https://www.whiteboxgeo.com/manual/wbt_book/available_tools/geomorphometric_analysis.html?highlight=topographic%20position#relativetopographicposition

filter size is in # of cells
```{r}
wbt_relative_topographic_position(
  dem = "hydem1m_smooth_filled_breached.tif", 
  output = "tpi100m.tif", 
  filterx = 100, 
  filtery = 100)

tpi100 <- raster("tpi100m.tif")

tmap_mode("view")
tm_shape(tpi100)+
  tm_raster(style = "cont",palette = "-Greys", legend.show = FALSE)+
  tm_scale_bar()

```

